Effects of pre-existing structures on the seismicity of the Charlevoix Seismic Zone

### 2019-06-29 main_full_manuscript.tex
saving progress

We conducted resolution tests in which the change in displacement and stress solutions relative to a 1 km-resolution model are monitored as a function of element size. \change[OF]{Displacement was almost constant for 2 km or smaller element sizes}{Displacements and differential stresses do not change appreciably for elements sizes equal to 4 km or smaller (Fig. S1)}. \add[OF]{Based on the resolution tests, we use the element size of 2 km for most of the models and 4 km for more computationally expensive models like those for exploring the effects of the friction coefficient. Low-friction models, especially those with $\mu$ equal to 0.1 and 0.2, are very expensive due to slow convergence rate in the non-linear solution scheme. Running all the models in this group at the same mesh resolution of 4 km ensures intra-group consistency; and the comparison with 2 km-resolution models is also justified because the descritization error does not dramatically increase as the resolution increases to 4 km (Fig. S1).} We also observe edge effects by comparing a 420 $\times$ 420 $\times$ 40 km model with the original 220 $\times$ 220 $\times$ 40 km model, finding that the seismogenic parts around the \change[OF]{crater}{impact structure} show consistent displacement and stress solutions for both domain sizes. Based on this result, we choose the smaller domain for this study and set the element size to be about 2 km. \add[OF]{The average computation time was about 16 days for fault models ($\mu$ = 0.3) with 2 km element size on 16 cores on the University of Memphis High Performance Computing cluster. The model with fault dips of 65$^\circ$-40$^\circ$-40$^\circ$ took about two times as long to run for the same model time. We do not clearly understand why the model with shallower-dipping faults is slower than the others roughly by a factor of two and further analyzing computational efficiency is beyond the scope of this paper.}
%we use an element size of 4 km for models with friction coefficient ($\mu$) values of 0.1 and 0.2 due to their expensive computation time.

@@

For simplicity, we assume that the \change[OF]{crater}{impact structure}'s bulk and shear moduli are a certain fraction of those of the crust~\citep[e.g.,][]{Baird_2010}. Similarly, the density of the \change[OF]{crater}{impact structure} is reduced \remove[OF]{by 10 \%} from that of the 1D velocity models over the corresponding depth range \add[OF]{to model a damaged crustal rock. Models with density reduced by 0 to 20 \% were run but did not exhibit significant changes as shown in Fig. S3 in the supplementary information. Unless stated otherwise, models presented in this study have 10 \% of density reduction in the impact structure.}

@@

\subsection{Treatment of rift faults}
We treat the rift faults as cohesive-frictional planes, on which slip vectors are not prescribed but determined by the model. In order to create relative motions across the rift fault planes, PyLith decouples the motion of the two sides of the faults by inserting cohesive cells on the fault surfaces. We assign a coefficient of static friction ($\mu$) and the cohesion ($C$) on the fault. We assume that $\mu$ and $C$ are uniform on the fault planes in each model but investigate the consequences of varying $\mu$ from 0.1 to 0.6 at a constant $C$ of 0 MPa. \add[OF]{We choose zero cohesion because it is appropriate for pre-existing faults with low normal stress}~\citep{Marone_Chris}. 
%\add[OF]{and slips on the model faults occur only down to about 10 km depth satisfying this condition.} 
%\note[EC]{I think the following sentences should be in discussion, not here.}\add[OF]{We ran a series of models in which the rift faults dip at 70$^{\circ}$ and the friction coefficient ($\mu$) is fixed to be 0.3 while cohesion ($C$) is one of 0, 3, 5, 10, 20 and 30 MPa (Fig. S4). Cohesion values less than 5 MPa could explain the spatial distribution of observed seismicity in the CSZ.}

@@

Magnitudes of $\Delta \sigma_{D}^{\prime}$ decrease as $\mu$ increases (Fig. \ref{fig:eff_of_mu}A) because the model behaviors must approach those of the no-fault model as fault strength increases. When $\mu$ values are 0.1 and 0.2, the area of positive $\Delta\sigma_{D}^{\prime}$ regions does not cover the entire observed seismicity\add[OF]{, with about 3\% and 24\% of the hypocenters in the region with $\Delta\sigma_{D}^{\prime} \ge$ 0.5\%, respectively}. The regions of positive $\Delta\sigma_{D}^{\prime}$ for $\mu$ values equal to 0.3 and 0.4 show better spatial correspondence with the seismicity of the CSZ. However, the model with $\mu = 0.4$ predicts positive Coulomb stress changes along the entire length of the rift faults, which is contrary to the limited spatial extent of the observed seismicity. \add[OF]{Also, about 66\% of the hypocenters in the model with a $\mu$ of 0.3 fall within the region of $\Delta\sigma_{D}^{\prime} \ge$ 0.5\%, compared to a value of 35\% when $\mu$ is 0.4. However, when we consider the region with $\Delta\sigma_{D}^{\prime} \ge$ 0\%, the percentage increases to 82\% when $\mu$ is 0.4 compared to 73\% in the model with $\mu$ equal 0.3.} $\Delta\sigma_{D}^{\prime}$ approaches zero when $\mu$ values are 0.5 and 0.6, for which the faults are locked. \add[OF]{The percentage of hypocenters in the region with $\Delta\sigma_{D}^{\prime} \ge$ 0.5\% reduces to 1.4\% and 0\% when $\mu$ values are 0.5 and 0.6, respectively.}

@@

\subsection{Frictional strengths of the rift faults}
The surface expression of the rift faults used in this study correspond to the Gouffre Northwest Fault, Saint-Laurent fault and Charlevoix Fault (Fig. 1). We did not consider the South Shore fault given its aseismic nature \change[OF]{and its location at the extreme eastern edge of the crater}{, although it appears to mark a boundary to the active zone}~\citep{lamontagne1999}. Our models show that a $\mu$ of 0.3 and cohesion values of 0 MPa create $\Delta\sigma_{D}^\prime$ maps that spatially correlate with the seismicity of CSZ and the observed stress rotations (Fig. \ref{fig:eff_of_mu}). \add[OF]{We ran a series of models in which the rift faults dip at 70$^{\circ}$ and the friction coefficient ($\mu$) is fixed to be 0.3 while cohesion ($C$) is one of 0, 3, 5, 10, 20 and 30 MPa (Fig. S6). Cohesion values less than 5 MPa could explain the spatial distribution of observed seismicity in the CSZ.} \add[OF]{A $\mu$ of 0.4 could also be suitable for the rift faults as the $\Delta\sigma_{D}^\prime$ maps spatially correlate with the seismicity of the CSZ. Some earthquakes also occurred at the SW region outside the impact structure (Fig. $\ref{figone}$), however, previous studies in the CSZ agreed with a finite extent of the seismicity}~\citep[e.g.,][]{Stevens1980}. 
\change[OF]{This}{Our fault model} is different from the work of \citet{Baird_2010} who modeled the Gouffre Northwest Fault, Saint-Laurent fault and South Shore Fault (Fig. \ref{figone}).

@@

\remove[OF]{In contrast to the normal faults as suggested by} 
%\citet{Hurd_Zoback_2012,Hurd_Zoback2012b,Fereidoni2014}, \citet{Baird_2010} 
\remove[OF]{conclude that the rift faults in the CSZ are very weak with a friction angle of 5$^\circ$ that is equivalent to a $\mu$ of about 0.1. This very low value of $\mu$ was necessary for their stopping criterion for boundary loading, which was $\sigma_{D}$ equal to 200 MPa at 10 km depth.} In our study, we stop increasing boundary displacements when $\sigma_{D}$ reaches 706 MPa at 10 km depth as described earlier without making any assumptions for the intermediate principal stress ($\sigma_2$). \add[OF]{In contrast,} \citet{Baird_2010} \add[OF]{used a $\sigma_{D}$ equal to 200 MPa at 10 km depth as their stopping criterion for boundary loading. They suggested that the rift faults in the CSZ are very weak with %a friction angle of 5$^\circ$ that is equivalent to 
a $\mu$ of about 0.1. The higher stopping criterion in our study is based on Byerlee's law under the assumption of no pore-fluid pressure on the rift faults, and it is necessary to make faults slip at seismogenic depth. Inclusion of pore-fluid pressure can significantly reduce the magnitude of $\sigma_{D}$ used as the stopping criterion.} The stopping criteria of a $\sigma_{D}$ of 200 MPa at 10 km depth would not make rifts faults with $\mu \ge 0.3$ slip \add[OF]{at seismogenic depths unless high pore fluid pressures are present. Another possible consequences of pore-fluid pressure is that faults could be reactivated at smaller angles from the maximum compressive stress}~\citep[e.g.,][]{ChenChen18}.

@@

Our models can only explain about 50\% of the observed $SH_{\max}$ rotation. A possible hypothesis for the remaining stress rotation is a change in the dip of the rift faults with depth and the effect of postglacial rebound. \remove[OF]{The high stress rotation could also be due to unconstrained focal mechanisms used in the stress inversion.} Most of the focal mechanisms used for stress inversion \add[OF]{in the CSZ} are obtained from first motion P and SH waves~\citep{Mazzotti_2010,lamontagne1998}. \change[OF]{Better constrained}{The} focal mechanisms \add[OF]{are good, and we suggest that the focal mechanisms can be further constrained} using waveform modeling \remove[OF]{will be needed} to validate the magnitude of the stress rotation.

>>>>

We conducted resolution tests in which the change in displacement and stress solutions relative to a 1 km-resolution model are monitored as a function of element size. \change[OF]{Displacement was almost constant for 2 km or smaller element sizes}{Displacements and differential stresses do not change appreciably for elements sizes equal to 4 km or smaller (Fig. S1)}. \add[OF]{Based on the resolution tests, we use the element size of 2 km for most of the models and 4 km for more computationally expensive models like those for exploring the effects of the friction coefficient. Low-friction models, especially those with $\mu$ equal to 0.1 and 0.2, are very expensive due to slow convergence rate in the non-linear solution scheme. Running all the models in this group at the same mesh resolution of 4 km ensures intra-group consistency; and the comparison with 2 km-resolution models is also justified because the descritization error does not significantly increase as the resolution increases to 4 km (Fig. S1).} We also observe edge effects by comparing a 420 $\times$ 420 $\times$ 40 km model with the original 220 $\times$ 220 $\times$ 40 km model, finding that the seismogenic parts around the \change[OF]{crater}{impact structure} show consistent displacement and stress solutions for both domain sizes. Based on this result, we choose the smaller domain for this study and set the element size to be about 2 km. \add[OF]{The average computation time was about 16 days for fault models ($\mu$ = 0.3) with 2 km element size on 16 cores on the University of Memphis High Performance Computing cluster. The model with fault dips of 65$^\circ$-40$^\circ$-40$^\circ$ took about two times as long to run for the same model time. We do not clearly understand why the model with shallower-dipping faults is slower than the others roughly by a factor of two and further analyzing computational efficiency is beyond the scope of this paper.}
%we use an element size of 4 km for models with friction coefficient ($\mu$) values of 0.1 and 0.2 due to their expensive computation time.

@@

For simplicity, we assume that the \change[OF]{crater}{impact structure}'s bulk and shear moduli are a certain fraction of those of the crust~\citep[e.g.,][]{Baird_2010}. Similarly, the density of the \change[OF]{crater}{impact structure} is reduced \remove[OF]{by 10 \%} from that of the 1D velocity models over the corresponding depth range \add[OF]{to model damaged crustal rocks. Models with density reduced by 0 to 20 \% were run but did not exhibit significant changes as shown in Fig. S3 in the supplementary information. All the models presented in this study have 10 \% of density reduction in the impact structure.}

@@

\subsection{Treatment of rift faults}
We treat the rift faults as cohesive-frictional planes, on which slip vectors are not prescribed but determined by the model. In order to create relative motions across the rift fault planes, PyLith decouples the motion of the two sides of the faults by inserting cohesive cells on the fault surfaces. We assign a coefficient of static friction ($\mu$) and the cohesion ($C$) on the fault. We assume that $\mu$ and $C$ are uniform on the fault planes in each model but investigate the consequences of varying $\mu$ from 0.1 to 0.6 at a constant $C$ of 0 MPa.  \add[OF]{Although cohesion values from 0 to 30 MPa have been tested and found to have significant effects, we choose to present models with zero cohesion in this study because it is appropriate for pre-existing faults with low normal stress}~\citep[e.g.,][]{Marone_Chris}. \add[OF]{We note that the deep portions of the model faults might not satisfy the condition of low normal stress and thus a further investigation on the depth-variation of cohesion is desirable.}
%\add[OF]{and slips on the model faults occur only down to about 10 km depth satisfying this condition.} 
%\note[EC]{I think the following sentences should be in discussion, not here.}\add[OF]{We ran a series of models in which the rift faults dip at 70$^{\circ}$ and the friction coefficient ($\mu$) is fixed to be 0.3 while cohesion ($C$) is one of 0, 3, 5, 10, 20 and 30 MPa (Fig. S4). Cohesion values less than 5 MPa could explain the spatial distribution of observed seismicity in the CSZ.}

@@

Magnitudes of $\Delta \sigma_{D}^{\prime}$ decrease as $\mu$ increases (Fig. \ref{fig:eff_of_mu}A) because the model behaviors must approach those of the no-fault model as fault strength increases. When $\mu$ values are 0.1 and 0.2, the area of positive $\Delta\sigma_{D}^{\prime}$ regions does not cover the entire observed seismicity\add[OF]{, with about 3 and 24\% of the hypocenters within the region of $\Delta\sigma_{D}^{\prime} \ge$ 0.5\%, respectively}. The regions of positive $\Delta\sigma_{D}^{\prime}$ for $\mu$ values equal to 0.3 and 0.4 show better spatial correspondence with the seismicity of the CSZ. However, the model with $\mu = 0.4$ predicts positive Coulomb stress changes along the entire length of the rift faults, which is contrary to the limited spatial extent of the observed seismicity. \add[OF]{Also, about 66\% of the hypocenters in the model with $\mu$ = 0.3 fall within the region of $\Delta\sigma_{D}^{\prime} \ge$ 0.5\% while only 35\% do when $\mu$ = 0.4. However, when the hypocenters falling in the region with $\Delta\sigma_{D}^{\prime} \ge$ 0\% are considered, the percentages increase to 73 and 82\% for $\mu$ equal to 0.3 and 0.4.} 
$\Delta\sigma_{D}^{\prime}$ approaches zero when $\mu$ values are 0.5 and 0.6, for which the faults are locked. \add[OF]{The percentage of hypocenters in the region with $\Delta\sigma_{D}^{\prime} \ge$ 0.5\% reduces to 1.4\% and 0\% when $\mu$ values are 0.5 and 0.6, respectively.}

@@

\subsection{Frictional strengths of the rift faults}
The surface expression of the rift faults used in this study correspond to the Gouffre Northwest Fault, Saint-Laurent fault and Charlevoix Fault (Fig. 1). We did not consider the South Shore fault given its aseismic nature \change[OF]{and its location at the extreme eastern edge of the crater}{, although it appears to mark a boundary to the active zone}~\citep{lamontagne1999}. Our models show that a $\mu$ of 0.3 and cohesion values of 0 MPa create $\Delta\sigma_{D}^\prime$ maps that spatially correlate with the seismicity of CSZ and the observed stress rotations (Fig. \ref{fig:eff_of_mu}). \add[OF]{We ran a series of models in which the rift faults dip at 70$^{\circ}$ and the friction coefficient ($\mu$) is fixed to be 0.3 while cohesion ($C$) is one of 0, 3, 5, 10, 20 and 30 MPa (Fig. S6). Cohesion values less than 5 MPa could better explain the spatial distribution of observed seismicity in the CSZ than the greater values.} \add[OF]{$\mu$ equal to 0.4 also seems suitable to the rift faults as the $\Delta\sigma_{D}^\prime$ maps spatially correlate well with the seismicity of the CSZ. A small number of earthquakes also occurred at the SW region outside the impact structure (Fig. $\ref{figone}$), however, previous studies in the CSZ agreed with a finite extent of the seismicity}~\citep[e.g.,][]{Stevens1980}. \note[EC]{I don't understand the point of the sentence about the finite extent of the seismicity. Is it necessary?}
\change[OF]{This}{Our fault model} is different from the work of \citet{Baird_2010} who modeled the Gouffre Northwest Fault, Saint-Laurent fault and South Shore Fault (Fig. \ref{figone}). \note[EC]{This sentence should be in the model setup unless you want to discuss implications of this difference.}

@@

\remove[OF]{In contrast to the normal faults as suggested by} 
%\citet{Hurd_Zoback_2012,Hurd_Zoback2012b,Fereidoni2014}, \citet{Baird_2010} 
\remove[OF]{conclude that the rift faults in the CSZ are very weak with a friction angle of 5$^\circ$ that is equivalent to a $\mu$ of about 0.1. This very low value of $\mu$ was necessary for their stopping criterion for boundary loading, which was $\sigma_{D}$ equal to 200 MPa at 10 km depth.} In our study, we stop increasing boundary displacements when $\sigma_{D}$ reaches 706 MPa at 10 km depth as described earlier without making any assumptions for the intermediate principal stress ($\sigma_2$). \add[OF]{In contrast,} \citet{Baird_2010} \add[OF]{used a $\sigma_{D}$ equal to 200 MPa at 10 km depth as their stopping criterion for boundary loading. They suggested that the rift faults in the CSZ are very weak with %a friction angle of 5$^\circ$ that is equivalent to 
a $\mu$ of about 0.1. The higher stopping criterion in our study is based on Byerlee's law under the assumption of no pore-fluid pressure on the rift faults, and it is necessary to make faults slip at seismogenic depth when $\mu$ is 0.3 as assumed in this study. Inclusion of pore-fluid pressure can significantly reduce the magnitude of $\sigma_{D}$ used as the stopping criterion. In other words, the lower stopping criterion, $\sigma_{D}$ equal to 200 MPa at 10 km depth, might be sufficient for rifts faults with $\mu$ = 0.3 or 0.4 to slip if pore fluid pressure is sufficiently high. Another possible consequences of pore-fluid pressure is that faults could be reactivated at smaller angles from the maximum compressive stress}~\citep[e.g.,][]{ChenChen18}.

@@

Our models can only explain about 50\% of the observed $SH_{\max}$ rotation. A possible hypothesis for the remaining stress rotation is a change in the dip of the rift faults with depth and the effect of postglacial rebound. \remove[OF]{The high stress rotation could also be due to unconstrained focal mechanisms used in the stress inversion.} Most of the focal mechanisms used for stress inversion \add[OF]{in the CSZ} are obtained from first motion P and SH waves~\citep{Mazzotti_2010,lamontagne1998}. \change[OF]{Better constrained}{The} focal mechanisms \note[EC]{What do you mean by ``focal mechanisms are `good\rq{}''? It would be good/nice to have better constrained ones, you mean?}\add[OF]{are good, and we suggest that the focal mechanisms can be further constrained} using waveform modeling \remove[OF]{will be needed} to validate the magnitude of the stress rotation.
